---
title: Hello world!
author: Jerid Francom
date: '2017-07-23'
slug: hello-world
summary: 'This is the first post in the `academic` theme for Hugo.'
categories:
  - R
tags:
  - R Markdown
draft: false
---

```{r setup, echo=FALSE, message=FALSE, warning=FALSE}
library(tidyverse)
library(DT) # for dynamic tables
```

This is the first post in the `academic` theme for Hugo. The scaffolding was created using the `Addin` menu dropdown and selecting `New Post`. This dialogue will provide the appropriate fields to set up a post or page. By default the subdirectory for posts is `post`. In the `academic` theme, other subdirectories are available for pages with particular content and content-specific attributes.

Here's a bit of code to show the `highlight_style` used here `github`. 

```{r}
x <- 1:10
paste("Number", x)
```

## Updating YAML metadata

You can, of course, update the YAML frontmatter directly, but for `categories` and `tags` you can use `blogdown` through the `Addin` menu to scan previously used categories and tags. To get this functionality select `Update Metadata` from the menu. 